---
output:
  github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# reumetsat

<!-- badges: start -->
[![CRAN version](https://www.r-pkg.org/badges/version-last-release/reumetsat)](https://cran.r-project.org/package=reumetsat) 
[![cran checks](https://badges.cranchecks.info/worst/reumetsat.svg)](https://cran.r-project.org/web/checks/check_results_reumetsat.html)
[![photobiologyInOut status badge](https://aphalo.r-universe.dev/badges/reumetsat)](https://aphalo.r-universe.dev/reumetsat)
[![R build status](https://github.com/aphalo/reumetsat/workflows/R-CMD-check/badge.svg)](https://github.com/aphalo/reumetsat/actions)
[![Documentation](https://img.shields.io/badge/documentation-reumetsat-informational.svg)](https://docs.r4photobiology.info/reumetsat/)
[![doi](https://img.shields.io/badge/doi-10.32614/CRAN.package.reumetsat-blue.svg)](https://doi.org/10.32614/CRAN.package.reumetsat)
<!-- badges: end -->

## Installation

Installation of the most recent stable version from CRAN:

```{r, eval=FALSE}
# install.packages("photobiologyInOut")
```

Installation of the current unstable version from R-Universe CRAN-like repository:

```{r, eval=FALSE}
install.packages('reumetsat', 
                 repos = c('https://aphalo.r-universe.dev', 
                           'https://cloud.r-project.org'))
```

Installation of the current unstable version from GitHub:

```{r, eval=FALSE}
# install.packages("remotes")
remotes::install_github("aphalo/reumetsat")
```

## Documentation 

HTML documentation is available at (https://docs.r4photobiology.info/reumetsat/), including a _User Guide_.

## Contributing

Pull requests, bug reports, and feature requests are welcome at (https://github.com/aphalo/reumetsat). Contribution of example data files that could
be supported in future versions will be very much appreciated.

## Citation

If you use this package to produce scientific or commercial publications, please cite according to:

```{r}
citation("reumetsat")
```

## License

© 2024 Pedro J. Aphalo (pedro.aphalo@helsinki.fi). Released under the GPL, version 2 or greater. This software carries no warranty of any kind.

