---
output:
  github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# surfaceuv

<!-- badges: start -->
[![CRAN version](https://www.r-pkg.org/badges/version-last-release/surfaceuv)](https://cran.r-project.org/package=surfaceuv) 
<!-- [![cran checks](https://badges.cranchecks.info/worst/surfaceuv.svg)](https://cran.r-project.org/web/checks/check_results_surfaceuv.html) -->
[![photobiologyInOut status badge](https://aphalo.r-universe.dev/badges/surfaceuv)](https://aphalo.r-universe.dev/surfaceuv)
[![R-CMD-check](https://github.com/aphalo/surfaceuv/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/aphalo/surfaceuv/actions/workflows/R-CMD-check.yaml)
[![Documentation](https://img.shields.io/badge/documentation-surfaceuv-informational.svg)](https://docs.r4photobiology.info/surfaceuv/)
<!-- badges: end -->
<!-- [![doi](https://img.shields.io/badge/doi-10.32614/CRAN.package.surfaceuv-blue.svg)](https://doi.org/10.32614/CRAN.package.surfaceuv) -->

## Development stage and plans

The main Surface UV data product supported is "Surface UV" from the AC SAF (atmospheric composition) project of EUMETSAT. In addition, "Surface UV" data from the OMI/Aura project hosted at NASA are supported giving access to additional variables. In each case, two file formats are supported. Currently, data are read much faster from HDF5 files than from the other formats.

An additional surface UV data source with broad, but not global, geographic coverage is the AC SAF high resolution Europe Surface UV product (EUV). Support of these date is planned for a future version of the package.

## Purpose

Surface UV irradiances and doses derived from remotely sensed data acquired with instruments on satellites are important as ground level measurements of these variables are rather sparse, with a much lower spatial resolution than temperature and rainfall. It is frequent to rely on remote sensing when studying effects of solar UV radiation on human, animal and plant health at a large scale. This package supports reading of data from files downloaded from open-access internet servers. These data have very few restrictions to their use, in most cases only the requirement to cite the source.

**O3M SAF Offline surface ultraviolet radiation product (OUV).** Data files from the Surface UV offline data product of AC SAF EUMETSAT can be downloaded from the server of the Finnish Meteorological Institute (FMI) [EUMETSAT AC SAF website](https://acsaf.org/). Two different data "ingestion" (`sUV_read_`) functions cater for two different types of files: HDF5 files with data on a geographic grid and one file per day, and text (.txt) files with time series data for a single geographic location.
For more information on these and other related data products, please, see the [EUMETSAT AC SAF website](https://acsaf.org/).

**OMI/Aura Surface UVB Irradiance and Erythemal Dose Daily L3 Global Gridded 1.0 degree x 1.0 degree V3 (OMUVBd).** The OMI/Aura Surface UV offline data are available through NASA and can be downloaded from the [NASA EARTHDATA website](https://disc.gsfc.nasa.gov/datasets/OMUVBd_003/summary?keywords=OMUVBd) as NetCDF4 files with the possibility of sub-setting before download. HDF5 files are also available for download without possibility of sub-setting. In both cases the data are provided as one file per day on a geographic grid basis. Two different data "ingestion" (`sUV_read_`) functions cater for the two different types of files. The OMI/Aura Surface UV offline data are available through NASA and can be downloaded from the [NASA EARTHDATA website](https://disc.gsfc.nasa.gov/datasets/OMUVBd_003/summary?keywords=OMUVBd). 

All functions can read one or more files in a single call. Irrespective of the number of files read, all `sUV_read_` functions return a single combined data frame object, with locations and time coordinates in long form. Additional query functions are provided to extract information from files, when possible, without reading them in whole.

## Installation

Installation of the most recent stable version from CRAN:

```{r, eval=FALSE}
# NOT YET IN CRAN
# install.packages("surfaceuv")
```

Installation of the current unstable version from R-Universe CRAN-like repository:

```{r, eval=FALSE}
install.packages('surfaceuv', 
                 repos = c('https://aphalo.r-universe.dev', 
                           'https://cloud.r-project.org'))
```

Installation of the current unstable version from GitHub:

```{r, eval=FALSE}
# install.packages("remotes")
remotes::install_github("aphalo/surfaceuv")
```

## Documentation 

HTML documentation is available at (https://docs.r4photobiology.info/surfaceuv/), including a _User Guide_ that is part of the package, and an on-line-only set of instructions on how to download files from the server of the Finnish Meteorological Institute (FMI) [EUMETSAT AC SAF website](https://acsaf.org/).

## Contributing

Pull requests, bug reports, and feature requests are welcome at (https://github.com/aphalo/surfaceuv). Contribution of example data files that could be supported in future versions will be very much appreciated.

## Citation

If you use this package to produce scientific or commercial publications, please cite according to:

```{r}
citation("surfaceuv")
```

## License

The package includes data example files redistributed in folder `extdata` that are © 2024 AC SAF project of EUMETSAT and used by courtesy of the AC SAF project of the EUMETSAT. If reused or redistributed a note similar to this must be included, but it is recommended to download them directly from the original source. 

The package includes data example files redistributed in folder `extdata` that contain modified Copernicus Sentinel data [2024]. If reused or redistributed a note similar to this must be included, but it is recommended to download them directly from the original source. 

Package code and documentation © 2024 Pedro J. Aphalo (pedro.aphalo@helsinki.fi). Released under the GPL, version 2 or greater. This software carries no warranty of any kind.

