---
output:
  github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# reumetsat

<!-- badges: start -->
[![CRAN version](https://www.r-pkg.org/badges/version-last-release/reumetsat)](https://cran.r-project.org/package=reumetsat) 
[![cran checks](https://badges.cranchecks.info/worst/reumetsat.svg)](https://cran.r-project.org/web/checks/check_results_reumetsat.html)
[![photobiologyInOut status badge](https://aphalo.r-universe.dev/badges/reumetsat)](https://aphalo.r-universe.dev/reumetsat)
[![R build status](https://github.com/aphalo/reumetsat/workflows/R-CMD-check/badge.svg)](https://github.com/aphalo/reumetsat/actions)
[![Documentation](https://img.shields.io/badge/documentation-reumetsat-informational.svg)](https://docs.r4photobiology.info/reumetsat/)
[![doi](https://img.shields.io/badge/doi-10.32614/CRAN.package.reumetsat-blue.svg)](https://doi.org/10.32614/CRAN.package.reumetsat)
[![R-CMD-check](https://github.com/aphalo/reumetsat/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/aphalo/reumetsat/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

## Purpose

Package 'reumetsat' currently only supports reading data files from the Surface UV offline data product of AC SAF EUMETSAT downloaded from the server of the Finnish Meteorological Institute (FMI). Two different functions cater for two different types of files: HDF5 files with data on a geographic grid and one file per day, and text (.txt) files with time series data for a single geographic location. Both functions can read one or more files. Irrespective of the number of files they return a single combined data frame object.

For more information of these and other data products, please, see the [EUMETSAT AC SAF website](https://acsaf.org/).

## Installation

Installation of the most recent stable version from CRAN:

```{r, eval=FALSE}
# install.packages("photobiologyInOut")
```

Installation of the current unstable version from R-Universe CRAN-like repository:

```{r, eval=FALSE}
# install.packages('reumetsat', 
#                  repos = c('https://aphalo.r-universe.dev', 
#                            'https://cloud.r-project.org'))
```

Installation of the current unstable version from GitHub:

```{r, eval=FALSE}
# install.packages("remotes")
remotes::install_github("aphalo/reumetsat")
```

## Documentation 

HTML documentation is available at (https://docs.r4photobiology.info/reumetsat/), including a _User Guide_.

## Contributing

Pull requests, bug reports, and feature requests are welcome at (https://github.com/aphalo/reumetsat). Contribution of example data files that could
be supported in future versions will be very much appreciated.

## Citation

If you use this package to produce scientific or commercial publications, please cite according to:

```{r}
citation("reumetsat")
```

## License

The data example files redistributed in 'extdata' are © 2024 AC SAF project of EUMETSAT and used by courtesy of the AC SAF project of the EUMETSAT. If reused of republished a note similar to this must be included. 

© 2024 Pedro J. Aphalo (pedro.aphalo@helsinki.fi). Released under the GPL, version 2 or greater. This software carries no warranty of any kind.

